Deep Learning does not Replace Bayesian Modeling

### 2021-10-16 Author Response Letter.Rmd

---
title: "Response to Reviewers for: Deep Learning does not Replace Bayesian Modeling"
author: "Breck Baldwin"
email: "breckbaldwin@gmail.com"
date: "10/15/2021"
output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html

bibliography: citations.bib
csl: american-medical-association.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	comment = NA
)
```

``` {echo=TRUE}
Reviewer: 1

Comments to the Author
An interesting paper with much to think on. It is noticeable that Deep Learning is now the chosen method in most fields even though it is computer science which dominates. This should perhaps be noted. It would be interesting to consider whether adopting Deep Learning in these fields is always appropriate or if Bayesian methods could offer a more appropriate alternative.
```

Author Response

R1:
This paper also demonstrates that deep learning is not the default tool when it is time for some reasoning in the scientific enterprise. If you are doing immunology/microbiology then Bayesian methods dominate. Environmental science, economics are a coin flip. While the choice of machine reasoning techniques should be entirely determined by the problem being addressed, researchers use what is popular and being talked about. Perhaps this paper gets Bayesian methods considered a bit more. 
:R1

Added comment

``` {echo=TRUE}
Reviewer: 2

Comments to the Author
The author provides a robust and necessary analysis of the interest around and funding dedicated to the areas of Bayesian inference and Deep Learning. The analysis yields interesting insights into the utilisation of these learning frameworks in a variety of fields. The article is concise, well written and I in my opinion is ready for publication as is.
```

``` {echo=TRUE}
Reviewer: 3

Comments to the Author
A thought-provoking paper, which with some refinement provides further evidence regarding the importance of computational statistics - often peddled as AI hype.

I have a series of questions & comments that I would welcome author response to.

Abstract: you comment “once computer science is subtracted”. How can you argue for this though, given that so much work in both deep learning and probabilistic modeling (though as below, please don’t assume these are different!) sits squarely in computer science! This feels, without full argument, like cherry-picking - and of course, in the subjects that are left (dominated by Stats & Applied Math) it’s inevitable that we find more usage of Bayesian methods (and probably more use of R as opposed to Python too!)

Google, Facebook, Microsoft and the gang are strong exponents of Bayes! Let’s remember that Bayesian Optimization sits at the core of hyper-parameter tuning for AlphaGoZero and (behind the scenes) for much model updating in Facebook.
```
### Abstract revised:

One could be excused for assuming that deep learning had or will soon usurp all credible work in reasoning, artificial intelligence and statistics, but like most 'meme' class broad generalizations the concept does not hold up to scrutiny. Memes don't generally matter since the experts will always know better but in the case of Bayesian software like Stan and PyMC3 even its developers and advocates bemoan the apparent dominance of deep learning as manifested in popular culture, breathtaking performance and most problematically from funding agency peer review that impacts our ability to further advance the field. The facts however do not support the assumed dominance of deep learning in science upon closer examination.

This letter simply makes the argument by the crudest of possible metrics, citation count, that once R2:the discipline of:R2 Computer Science is subtracted, Bayesian software accounts for nearly a third of research citations. Stan and PyMC3 dominate some fields, PyTorch, Keras and TensorFlow dominate others with lots of variation in between. Bayesian and deep learning approaches are related but very different technologies in goals, implementation and applicability with little actual overlap so this is not a surprise, R3: e.g., deep learning cannot bring the explainability of applied math/statistics and Bayesian methods do not scale to deep learning data sets.

While deep learning behemoths like Facebook and Google both use and support Bayesian efforts, the Bayesian packages scientists actually use are academic/volunteer efforts punching far above their weight class and they need financial support. It would behoove funders to fully understand the impact and role of Bayesian methods in resource allocation.:R3

```{echo=TRUE}

Introduction: There is a *huge* rise in the use of Bayesian Deep learning. Please do not fall into the trap of making DL look like some ‘other’ thing. It’s just a non/hyper-parametric model and as such we can do all the full Bayesian stuff - and many do! There are well-documented works on semi-structured HMC, Riemann MC and good old fashioned MCMC for use in big deep nets, let alone all the neat work on approximate Bayes, mainly using variational learning. Look at work on things like loss-calibrated Bayesian deep nets, Bayes DL, Bayesian autoencoding, ‘HamilTorch’ MCMC package for TensorFlow (https://adamcobb.github.io/journal/hamiltorch.html) etc. You just can’t ignore all this. (and please don’t forget all the fine folks doing {S,O,P}DiffEq learning using things like graph NNs and ResNets - lots of Bayes to please everyone in the ‘physics meets DL’ world).
```

Intro: you claim NSF dismissal of (eg) Stan - where is the evidence citation for this?

Intro: Bayesian software - myopic - more to Bayes than sampling (which is non-Bayesian in any case!). Lots of work in AI using approximate Bayes, ABC, VB and so on. Gaussian Processes are HUGE in their own right too - but no MCMC in sight normally. You really need to note that Gaussian Processes, and Neural Kernel methods, are *fully Bayesian* yet DL replacements for many applications.

Sect 3. I just do not buy that packages are a proxy for research work. As before, what about all the "Bayesian" dropout work in NNs etc. VAEs? BayesDL tends not to use packages.

So, this has the making of a really interesting commentary - but really needs to address these issues before it can be published. I feel you do Bayes in DL a disservice by commenting just on big package usage.
```

Reviewer: 4

Comments to the Author
The author presents a lightweight but thought-provoking exposition of the scientific literature, providing interesting evidence that can be used by grantees and funders alike to justify funding statistical science (in particular Bayesian statistics) alongside broader computer science techniques badged under "AI". I was slightly surprised not to see references to Bayesian deep learning, or work at the intersection of Bayesian statistics and deep learning (e.g. using DNN as an approximation to a likelihood function), to support the point around complementarity, but this is a minor point and not worth a revision to the paper. I do hope the paper provokes further debate on this important topic.

#

### 2021-10-16 DeepLearningAndBayesianSoftware_changes_noted.Rmd
backup commit, drafts not ready

1. How does Bayesian modeling software stack up against deep learning without appeal to feature comparisons, performance arguments on suspect data sets or achieved closeness to Platonic ideals? Just go out and count how many citations the respective approaches have in the research literature. Counting and categorization, that's it. 
2. Asses the impact of Bayesian modeling software using deep learning metrics as a yard stick--how big a fraction of a huge thing are we?
3. Contextualize the roles each approach has by looking at subject distributions. The technologies have very different use cases so one would expect variation.

@@

@@

>>>>

1. How does Bayesian modeling software stack up against deep learning without appeal to feature comparisons, performance arguments on suspect data sets or achieved closeness to Platonic ideals? My response is--just go out and count how many citations the respective approaches have in the research literature. Counting and categorization, that's it. 
2. Asses the impact of Bayesian modeling software using deep learning metrics as a yard stick--how big a fraction of a huge thing are we?
3. Contextualize the roles each approach has by looking at subject distributions. The technologies have very different use cases so one would expect variation.

@@

R4: 
## Characterizing the difference between Stan/PyMC3 and TensorFlow/PyTorch

What, if anything, are the hard differences between Bayesian modeling and deep learning? Deep learning can be implemented in Bayesian modeling, and lots of Bayesian concepts get used in the deep learning world as one reviewer noted:

"There is a *huge* rise in the use of Bayesian Deep learning. Please do not fall into the trap of making DL look like some ‘other’ thing. It’s just a non/hyper-parametric model and as such we can do all the full Bayesian stuff - and many do! There are well-documented works on semi-structured HMC, Riemann MC and good old fashioned MCMC for use in big deep nets, let alone all the neat work on approximate Bayes, mainly using variational learning. Look at work on things like loss-calibrated Bayesian deep nets, Bayes DL, Bayesian autoencoding, ‘HamilTorch’ MCMC package for TensorFlow (https://adamcobb.github.io/journal/hamiltorch.html) etc."

Perhaps Stan and PyMC3 are really part of the same discipline as deep learning and this article catalogs usage differences of siblings from the same parents--sort of like the difference between R and Python?

But no. Another reviewer comment contains the key insight: they said it was no surprise that Bayesian software had high usage outside of computer science since other subjects are "dominated by Stats & Applied Math". I took this to mean that deep learning is not a natural candidate for use in those fields, and software like Stan and PyMC3 are. Unpacking that a bit I'll observe:

* Unless the phenomenon under study is a physical neural net, deep learning only offers prediction services. While outstanding progress has been made and further progress may well involve Bayesian concepts, it is only prediction in the end. 
* The basis of prediction in deep learning is opaque to human comprehension even if Bayesian techniques are used. This applies to generative neural nets as well. 
* Opacity blocks use in fields "dominated by Stats & Applied Math" where the goal involves developing and fitting mechanistic models. The science is in the model description, quality of fit validates the model. A high quality fit in the absence of an understandable model does not help.

Deep learning clearly is used in mechanistic models but typically as a sensor or classifier, e.g., classify x-ray images for evidence of COVID pneumonia in an epidemialogical study. That study itself will likely be a state-based model where transition rates are explicitly estimated statistically and are human interpretable.

Another case are hybrid models where a deep learning component replaces some or all of the likelihood of a Bayesian model. While the overall model may be interpretable, the individual deep learning components remain black boxes. Interpretability comes from have well understood properties from how they were trained etc. For example Maggie Lieu's presentation at StanCon that used a deep learning component to substitute for a numerically unstable galaxy mass estimator().

Also Bayesian models can be utterly opaque if not authored with an eye to understandability, but the option for understandability exists and is generally the expectation. Deep learning systems cannot match this because a big portion of 'authoring' (hyper parameter setting) in those frameworks are beyond human understanding.

In the game of "one of these things is not like the other", human interpretable models distinguishes Stan and PyMC3 from PyTorch and TensorFlow in quite absolute terms[^3].

If I am correct that deep learning and Bayesian modeling are so different when it comes to interpretablity then the mingling of the fields baffles the mind. They share similar programming techniques, have similar research leading edges and ... . 
I think Stan got a hold of the AI tail which whipped it out of booring stats into ML.

[^3]:However, in a year this article may be much more difficult to write since both projects have adopted HMC implementations but I'd suggest that is more of a grafting of another approach to their code base rather than natural growth of deep learning techniques into the PyMC3/Stan space.

:R4

## Defining what counts as a citation

@@

R1: This paper also demonstrates that deep learning is not neccessarily the default tool when it is time for some reasoning in the scientific enterprise. If you are doing immunology/microbiology then Bayesian methods dominate. Environmental science, economics are a coin flip. Researchers use what is fashionable perhaps as a default--it may ease publication, be more familiar to students/postdocs but at the potential cost of being a bad fit for the problem at hand. If Bayesian methods are good enough for immunologists then perhaps other fields should have a look.:R1
